\chapter{An Alternative Language for Describing Type Systems}

Syntactic constructs are build in stages.

Firstly, for each construct we define the introduction form and
assocated type \emph{checking} rule.

An optional rule may be provided to judge when the construct is a
\emph{type} (of any universe). 

We then define a syntax for elimination using the symbols $<construct>$
and $<eliminator>$ (although not every $<eliminator>$ need to contain
another language construct, it may be an atom (like $proj_l$ and
pair), internally all elimination forms are represented as
pairs. Elimination is optional and need not be provided, and we can
also supply more than one elimination for per construct. We must be
able to synthesize a type for $<construct>$ therefore rather than
being the raw introduction form, it is either another (synthesizable)
computation, or an introduction form paired with a type
annotation. Only a maximum of one $<eliminator>$ is allowed, however
this can be of arbirary complexity, i.e. in the rule you could specify
that the eliminator be a list of things.

\cite{TypesWhoSayNi} page 18 - we are free to write synthesis rules
ONLY for eliminators, we must name the target but not peek at its
structure. The first premise of the elimination rule must synthesize
the target's type. So what information do we really need for our
elimination rules internally? Well, we will need to give a pattern
that we wish to match against the target's type synthesis, we will
need to specify the type at which we are checking the $<eliminator>$
and finally the synthesized type we are returning. The target in an
elimination must be \emph{radical} - i.e be a synthesizable term
(computation) or a construction with a type annotation.

We will need to supply a $\beta$ rule for each elimination we
give. Actually in \cite{TypesWhoSayNi} page 18/19, for the beta rule,
it is almost enough to just the term it reduces to, we can work out
much of the beta rule information from the typing rules!

Type checking may be required to validate user-given $\beta$ rules!
Specifically, we may need to employ previously defined $\beta$ rules
and allow some computation.

In \cite{TypesWhoSayNi} all $\beta$-rules have form:
$$(t:T)\;s \rightsquigarrow_{\beta} (t' : T')$$
and t, T and s can demand only essential construction structure

A type synthesis rule for the elimination form must also be provided.

If an elimination form is provided, small step semantics must also be
provided (reduction to weak-head-normal form).

\section{Thoughts on check vs synthesize}

What is checkable?

\begin{itemize}
\item Introduction forms
\item Constants
\item Anything synthesizable
\end{itemize}

What is synthesizable?

\begin{itemize}
\item Variables
\item \emph{Valid} elimination forms (where eliminatee is synthesizable)
\item Anything checkable if it is annotated
\end{itemize}
