\chapter{Building a type checker generator}

Now that a user has the means of describing a type system, we will now
discuss how the generic type-checking software is constructed.

We will describe key definitions and types in order to communicate a
high level understanding of how the software is designed to work,
however we will often omit implmentation details in areas where the
code is less important, tedius or verbose.

The general tactic for type-checking some generic syntax will be to
parse the required typing, $\beta$ and $\eta$ rules from the
descriptions provided using our DSL. Using this information we can
then parse user source code and represent it in a fairly language
agnostic interal object syntax before using the rules to type-check
our internal syntax.

We will begin by describing the internal syntax, talk about how we
might represent the various tools and rules that we need in order to
type check this syntax and define the type checking process before
breifly covering our approach to parsing this information from the
user supplied input files. We purposely keep our parsing conversations
brief to prioritize the presentation of challanges in other areas.

Our approach in this section is heavily influenced by the work of
Conor McBride in \emph{The types who say 'ni'} \cite{TypesWhoSayNi} -
although the code we provide is strictly our own in all cases, there
are certainly areas where this code is a near direct translation of
the mathematics in this paper. As we pointed out in the previous
chapter: patterns, expressions and and the treatment of premises are
all concepts that we take from this work. In addition our internal
syntax is very like that described in the paper and we use the same
approach for thinnings, substitution and contexts.

\\\\
\input{../../TypeCheckerGenerator/latex/CoreLanguage.tex}
\input{../../TypeCheckerGenerator/latex/Thinning.tex}
\input{../../TypeCheckerGenerator/latex/Substitution.tex}
\input{../../TypeCheckerGenerator/latex/TermSubstitution.tex}
\input{../../TypeCheckerGenerator/latex/Context.tex}
\input{../../TypeCheckerGenerator/latex/Opening.tex}
\input{../../TypeCheckerGenerator/latex/Pattern.tex}
\input{../../TypeCheckerGenerator/latex/Expression.tex}
\input{../../TypeCheckerGenerator/latex/Rules.tex}
\input{../../TypeCheckerGenerator/latex/EtaRule.tex}
\input{../../TypeCheckerGenerator/latex/Semantics.tex}
\input{../../TypeCheckerGenerator/latex/TypeChecker.tex}
\input{../../TypeCheckerGenerator/latex/Parser.tex}
\input{../../TypeCheckerGenerator/latex/SpecParser.tex}
\input{../../TypeCheckerGenerator/latex/LanguageParser.tex}

